Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-02-22 paper/manuscript.Rmd
header

# Introduction
Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, the energetic
expenditure of swimming for quadrupedal mammals is significantly higher than
walking or running <!--[@Fish1993]-->, while drowning is also possible [@Miller1985].
Despite a large number of coastal and island caribou herds, only a handful of
examples have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012].

@@

# Forage limitation
Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer2016] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer2016]. Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller1977]. For very small islands, over-grazing by newly
arrived caribou may rapidly deplete forage [@Bergerud1990]. Without sea ice,
movement from the mainland to an island, or between islands, requires caribou to
assess the trade-off associated with swimming.

@@

# Predator avoidance
Predator avoidance could also contribute to swimming behaviour of caribou
[@Miller2002]. During calving, female caribou are known to swim to small islands
in freshwater lakes and give birth, or with their calves after birth, to avoid
predation [@Bergerud1985]. The use of shoreline habitat in summer is considered
an effective anti-predator calving strategy by allowing for swimming to islands
or peninsulas as a predation avoidance flight response [@Bergerud1990]. The use
of water, and by association occasionally swimming, is therefore part of the
fine-scale interactions between caribou and their predators and the use of
islands appears to be an effective anti-predator strategy.

Historic predators of caribou such as wolves (*Canis lupus*) and black bears
(*Ursus americanus*) are not present in the Fogo Island archipelago, thus
leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
adult caribou is relatively rare [@BastilleRousseau2016], so the potential
effect of predators on caribou swimming behaviour is limited to non-consumptive
effects associated with encounters and subsequent flight response. Adult females
with calves at heel may be an exception because calves are highly susceptible to
coyotes [@Rayl2014], but the energetic costs for calves to swim in the ocean is
likely far greater than adults. As one might expect, none <!--confirm--> of the
86 swimming events we observed occurred in the weeks after calves were born,
suggesting swimming with in the ocean with a calf-at-heel is likely more risky
than other predator avoidance behaviours.

@@

# Individual variation 
In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--double check--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. <!-- QW: not sure how to finish this off -->
<!-- QW: I'm not sure we need to say this.. 
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

@@

# Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation to a series of proximate observations, fine-scale density-dependent
habitat selection is theoretically possible [@Webber2018]. Predator avoidance
could also contribute to caribou swimming behaviour, but given the presumed low
density of coyotes on Fogo Island and that predation primarily only affects
young calves, we suspect predator risk is a possible, but less likely
explanation for swimming.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
QW: changed it a bit - how is this? -->

@@

# Acknowledgements 
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided by L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
McClement, and S. Mahoney for data collection, management, and logistics. We
thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
on caribou in Newfoundland. Funding for this study was provided by the National
Sciences and Engineering Research Council (QMRW, JGH, EVW). We respectfully
acknowledge the territory in which data were collected and analyzed as the
ancestral homelands of the Beothuk, and the island of Newfoundland as the
ancestral homelands of the Mi’kmaq and Beothuk.
<!-- Simplified the funding statement, do we need to name the specific scholarships/grants? Eric opinion??-->

>>>>

**Introduction**

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, the energetic
expenditure of swimming for quadrupedal mammals is significantly higher than
walking or running <!--[@Fish1993]-->, while drowning is also possible [@Miller1985].
Despite a large number of coastal and island caribou herds, only a handful of
examples have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012].

@@

**Forage limitation**

Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer2016] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer2016]. Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller1977]. For very small islands, over-grazing by newly
arrived caribou may rapidly deplete forage [@Bergerud1990]. Without sea ice,
movement from the mainland to an island, or between islands, requires caribou to
assess the trade-off associated with swimming.

@@

**Predator avoidance**

Predator avoidance could also contribute to swimming behaviour of caribou
[@Miller2002]. During calving, female caribou are known to swim to small islands
in freshwater lakes and give birth, or with their calves after birth, to avoid
predation [@Bergerud1985]. The use of shoreline habitat in summer is considered
an effective anti-predator calving strategy by allowing for swimming to islands
or peninsulas as a predation avoidance flight response [@Bergerud1990]. The use
of water, and by association occasionally swimming, is therefore part of the
fine-scale interactions between caribou and their predators and the use of
islands appears to be an effective anti-predator strategy.

Historic predators of caribou such as wolves (*Canis lupus*) and black bears
(*Ursus americanus*) are not present in the Fogo Island archipelago, thus
leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
adult caribou is relatively rare [@Bastille-Rousseau2016], so the potential
effect of predators on caribou swimming behaviour is limited to non-consumptive
effects associated with encounters and subsequent flight response. Adult females
with calves at heel may be an exception because calves are highly susceptible to
coyotes [@Rayl2014], but the energetic costs for calves to swim in the ocean is
likely far greater than adults. As one might expect, none <!--confirm--> of the
86 swimming events we observed occurred in the weeks after calves were born,
suggesting swimming with in the ocean with a calf-at-heel is likely more risky
than other predator avoidance behaviours.

@@

**Individual variation**

In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--double check--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. <!-- QW: not sure how to finish this off -->
<!-- QW: I'm not sure we need to say this.. 
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

@@

**Summary and conclusion**

For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation to a series of proximate observations, fine-scale density-dependent
habitat selection is theoretically possible [@Webber2018]. Predator avoidance
could also contribute to caribou swimming behaviour, but given the presumed low
density of coyotes on Fogo Island and that predation primarily only affects
young calves, we suspect predator risk is a possible, but less likely
explanation for swimming.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
QW: changed it a bit - how is this? -->

@@

**Acknowledgements**

We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided by L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, C. Doucet, W.
Barney, and J. Neville for logistical support in the field as well as help with
data data collection and management. We thank T. Bergerud and S. Mahoney for
their vision in initiating much of the work on caribou in Newfoundland. Funding
for this study was provided by the National Sciences and Engineering Research
Council (QMRW, JGH, EVW). We respectfully acknowledge the territory in which
data were collected and analyzed as the ancestral homelands of the Beothuk, and
the island of Newfoundland as the ancestral homelands of the Mi’kmaq and
Beothuk.
<!-- Simplified the funding statement, do we need to name the specific scholarships/grants? Eric opinion??-->

\captionsetup[figure]{labelformat=empty}

![**Figure 1**: Image of swimming unmarled adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.](../graphics/IMG_1562.JPG)
